Binning Strategies and Related Loss for Binned Scatterplots of Large Data

### 2016-12-03 AuthorResponseToReviewers.Rmd

*For the other ideas covered in these two sections, please properly cite relevant work.*

- You are correct that many of the ideas in Sections 2 and 3 are not original, with the exception of the random binning approach, where the methodology for random-binned aggregation is original to my best knowledge. The goal of these sections is to bring many binning related concepts under a unified notation for the benefit of clarity in the sections to follow. For the random binning, a statement has been added to address the motivation for its introduction: deterministic one-sided interval bins lead to lopsided allocation of points on bin boundaries, whereas a random assignment is expected to split the points evenly across bins. For the unoriginal ideas, additional citations have been made to place the unifying notation better within the existing literature.

*Unconvincing Results:*
*In Section 4.1, you propose the idea of spatial loss as a metric to account for the loss of information. Please explain how this approach is related to Scott’s approach of using mean integrated square error. What makes your definition a better choice?*

- The comparison and benefit over Scotts MISE is not included to the discussion leading into Section 4.

@@

*I do not understand how binned frequencies (as compared to true frequencies) are calculated in Section 4.2 on frequency loss. Please provide an example. Also please clarify is binned frequencies are measures of frequencies perceived by a human.*

- A clarifying example using the visual from Figure 3 has been added to Section 4.2.

@@

*The first paragraph of Section 5.1 shows that spatial loss increases as bin size increases, this seems hardly surprising. In the second paragraph, you talk about three properties of visual estimator: unbiasedness, consistency and efficiency, I think these properties are interesting, but the rest of the paper does not address them sufficiently.*

*I also fail to understand how the discussion on artificial striping in Section 5.1 is relevant for spatial loss.*

- Artificial striping is a symptom of poorly aligned bin sizes and the data resolution, which will lead to imbalanced bin allocation and thus more spatial loss. A clarification of this idea and a visual comparison of bin sizes and spatial loss have been added to section 5.1 in the context of Figure 6.

### 2016-12-03 versionForJCGS/PaperJCGS.Rnw
Update text to address addition of Figure 6(b). Also address reviewer in doc.

To demonstrate the importance of properly selecting binwidths we consider the coarse version of the simulated bivariate-uniform data which is recorded to a resolution of two units in each dimension. Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios. Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal artificial stripes with five-by-five unit bins, and the appropriate view of an evenly spread density with four-by-four unit bins.  Note that random binning is effective at smoothing out the artificial striping patterns when non-integer multiples of the data resolution are selected.

@@

ggplot(aes(x=binxs, y=binys), data=allbindat) +
  geom_tile(aes(fill=binfreq,height=height, width=width), data=allbindat)+
  facet_grid(BinningType~BinSize, scales="free", space="free")+
  ggtitle(" ") +   xlab("") + ylab("") +
  scale_fill_gradient(low="#56B1F7", high="#132B43",
                      limits=c(0, 450),name="Bin Count") + 
  theme_bw() + theme(plot.margin = unit(c(0.1,0.1,-0.5, 0.1),"cm"),
                     aspect.ratio=1,
                     panel.grid= element_blank()) +
  theme(legend.position="bottom")

@@

\begin{figure}[hbtp]
\centering
  \subfloat[Binned scatterplots]{\includegraphics[keepaspectratio=TRUE,width=.66\textwidth]{./figure/CoarseUnif.png}}
  \subfloat[Spatial loss vs bin sizes]{\includegraphics[keepaspectratio=TRUE,width=.32\textwidth]{./figure/CoarseUnifLosses.png}} 
  \caption[Traditional and adapted scatterplots for games vs. strikeouts data.]{\label{fig:UniformStripes} Binned scatterplots and spatial losses for various sized bins with coarse uniform data. Note the slight dip in loss at even integers, when bins are aligned with data resolution.}}
\end{figure}

@@

\ktm{For the baseball data this leads us to specify small standard rectangular bins for data that is recorded with a one game by one strike-out data resolution. The data has only 42,583 observations, so we can use minimal binning to match the data resolution without taking an inordinate amount of computation time. While minimal binning eliminates spatial loss, the miniscule physical size of the resulting tiles makes it difficult to see the frequency shade; Therefore slightly larger two game by ten strikout bins are used, providing approximately 50 bins per dimension. The origin offset of $(0.5,-0.5)$ is used to shift the binning back by a half game and down by one half strikeout to adjust for the data resolution. This offset for reduces the spatial loss by approximately 8\% for this bin size.}

>>>>

To demonstrate the importance of properly selecting binwidths we consider the coarse version of the simulated bivariate-uniform data which is recorded to a resolution of two units in each dimension \ktm{-- thus even integer sized bins are most desireable. Figure~\ref{fig:UniformStripes}(a) displays the binned scatterplots under several scenarios. Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal artificial stripes with five-by-five unit bins, and the appropriate view of an evenly spread density with four-by-four unit bins. Figure~\ref{fig:UniformStripes}(b) demonstates that with coarse data resolution there are departures from the linear relationship between spatial loss and bin sizes; with small spikes in spatial loss where the bin size is highly misaligned with the data resolution and slight drops in spatial loss when aligned perfectly. In this we can see that artificial striping is a symptom of poorly aligned bin specification.}

@@

ggplot(aes(x=binxs, y=binys), data=allbindat) +
  geom_tile(aes(fill=binfreq,height=height, width=width), data=allbindat)+
  facet_grid(BinningType~BinSize, scales="free", space="free")+
  ggtitle(" ") +   xlab("") + ylab("") +
  scale_fill_gradient(low="#56B1F7", high="#132B43",
                      limits=c(0, 450),name="Bin \n Count") + 
  theme_bw() + theme(plot.margin = unit(c(0.1,0.1,-0.5, 0.1),"cm"),
                     aspect.ratio=1,
                     panel.grid= element_blank())

@@

\begin{figure}[hbtp]
\centering
  \subfloat[Binned scatterplots]{\includegraphics[keepaspectratio=TRUE,width=.63\textwidth]{./figure/CoarseUnif.png}} 
  \hspace{.05\textwidth}
  \subfloat[Spatial loss vs bin sizes]{\includegraphics[keepaspectratio=TRUE,width=.31\textwidth]{./figure/CoarseUnifLosses.png}} 
  \caption[Traditional and adapted scatterplots for games vs. strikeouts data.]{\label{fig:UniformStripes} \ktm{Binned scatterplots and spatial losses for various sized bins with coarse uniform data. Note the slight dip in loss at even integers, when bins are aligned with data resolution.}}
\end{figure}

Note that random binning is effective at smoothing out the artificial striping patterns \ktm{when many point fell along bin boundaries -- accomplishing the motiviation for its development. However due to inflated spatial loss due to the stocastic bin assignment, standard binning with properly selected bin dimension is the better option}.

@@

\ktm{For the baseball data this leads us to specify small standard rectangular bins for data that is recorded with a one game by one strike-out data resolution. The data has only 42,583 observations, so we can use minimal binning to match the data resolution without taking an inordinate amount of computation time. While minimal binning eliminates spatial loss, the minuscule physical size of the resulting tiles makes it difficult to see the frequency shade; Therefore slightly larger two game by ten strikeout bins are used, providing approximately 50 bins per dimension. The origin offset of $(0.5,-0.5)$ is used to shift the binning back by a half game and down by one half strikeout to adjust for the data resolution. This offset for reduces the spatial loss by approximately 8\% for this bin size.}
